Theory of non-dilute binding and surface phase separation applied to membrane-binding proteins

Version 1

Abstract

Abstract

Surface binding and surface phase separation of cytosolic scaffold proteins on lipid membranes are involved in many cellular processes, such as cell signaling, cell adhesion, and cortex regulation. However, the interplay between surface binding and surface phase separation is poorly understood. In this work, we study this interplay by deriving a general thermodynamic model and applying it to in vitro reconstitution experiments of membrane-binding proteins involved in tight junction initiation. Our theory extends the classical surface binding isotherm to account for non-dilute and heterogeneous conditions where components can phase separate. We use our theory to demonstrate how surface phase separation is governed by the interaction strength among membrane-bound scaffold proteins and their binding affinity to the membrane surface. Comparing the theory to reconstitution experiments, we show that tuning the oligomerization state of the adhesion receptors in the membrane controls surface phase transition and patterning of the scaffold protein ZO1. These findings suggest a fundamental role of the interplay between non-dilute surface binding and surface phase separation in forming the tight junction. More broadly, our work highlights non-dilute surface binding and surface phase separation as a common organizational principle for membrane-associated structures in living cells.

Introduction

Liquid-liquid phase separation (LLPS) has emerged as a fundamental principle governing the spatial organization of cellular components into dynamic, membrane-less compartments termed biomolecular condensates Brangwynne et al. (2011); Mahen and Venkitaraman (2012); Gall (2003); Brang-wynne et al. (2009); Fritsch et al. (2021); JR and R. (2009); CJ and R. (2012). Such condensates create distinct microenvironments that influence a wide range of cellular functions, including signaling Su et al. (2016); Wang et al. (2021); Xiao et al. (2022), transcription Boija et al. (2018); Lu et al. (2020), and stress responses Riback et al. (2017).

While LLPS has been extensively studied in the cytoplasm, protein phase separation is also involved in many biological processes happening at membrane surfaces such as signaling Li et al. (2012); CASE et al. (2019); Zeng et al. (2019); Li et al. (2022); Zeng et al. (2018); Su et al. (2016); Zheng et al. (2022), synapse formation Zeng et al. (2018), cell adhesion Beutel et al. (2019); Pombo-García et al. (2024); Wang et al. (2021) and cortex regulation CASE et al. (2019); Wiegand and Hyman (2020). Protein condensation at membranes is often induced by binding of multivalent scaffold proteins present in the cytoplasm to integral membrane proteins or lipids on the surface of the membrane. Membrane binding of scaffold proteins can result in condensation far below the concentration required for phase separation in the bulk, providing spatial-temporal control over the condensation process. Therefore, membrane binding and surface phase separation are tightly linked processes GrandPre et al. (2024).

How membrane binding and surface phase separation influence each other is not well understood. Classical surface binding theories, such as the Langmuir isotherm Langmuir (1918), assume dilute concentrations where binding sites are independent and non-interacting. These assumptions are not valid for phase separating systems, where surface densities are non-dilute and inter-molecular interactions among membrane-bound molecules become important. More recently, non-dilute binding and phase separation have been incorporated into theoretical models Zhao et al. (2021, 2024); Liese et al. (2023). However, these models still rely on simplifying assumptions such as uniform binding sites, which is not matching the reality of biological membranes.

Binding of proteins to biological membranes is typically tightly controlled in space and time by either increasing local densities of binding sites such as in the formation of cell adhesion complexes Otani et al. (2019); Zhao et al. (2018b), or by increasing binding affinities for example by phosphorylation Li et al. (2012). To accurately describe such phenomena, theoretical models must incorporate both non-homogeneous and non-dilute surface densities of proteins at the membrane. While a recent thermodynamic model has linked binding heterogeneity to prewetting phenomena Rouches et al. (2021), developing a theory that integrates these heterogeneous, non-dilute conditions and captures the potential for surface-bound phase transitions remains crucial for a comprehensive understanding of membrane-associated assemblies.

In this work, we present a theoretical framework for the interplay between surface binding and surface phase separation that includes heterogeneous and non-dilute membrane surface densities. We show how the classical binding isotherm must be fundamentally modified when molecules bound to membrane receptors interact strongly and undergo surface condensation. We derived a generalized binding isotherm which extends classical adsorption models by accounting for the non-dilute nature of surface-bound complexes and their interactions. Using this framework, we explored the behaviors of non-dilute surface binding under varying bulk and surface compositions, binding affinities, and interaction strengths between membrane bound complexes. To illustrate the applicability of our theory, we focus on in vitro reconstituted membrane systems that model components of tight junctions. We explain how oligomerization states influence membrane organization. More generally, our theory reveals how non-dilute surface binding and membrane-associated phase separation interact to influence membrane organization, shedding light on a fundamental physicochemical mechanism through which cells regulate membrane associated assembly processes.

Model for protein binding to receptors on a membrane

In the first part of this work, we discuss a theoretical framework that describes binding of soluble scaffold proteins to lipid membranes via interactions with receptor proteins embedded in a membrane. In addition, the scaffold proteins can undergo (surface) phase separation, which results in formation of protein-dense and protein-dilute regions on the membrane (Fig. 1a,b). The theoretical framework builds upon previous works on the interplay of surface-associated phase transition (surface phase separation, prewetting) and homogeneous surface binding Zhao et al. (2021, 2024). However, previous models do not involve a receptor component, reducing the stable coexistence of domains poor and rich in bound protein to a specific value of the bulk protein concentration where the phase transition occurs. However, this behavior is not consistent with in vitro experiments where phase coexistence exists over a large range of bulk protein concentrations. In this light, we propose an extended but still minimal model in terms of the number of membrane-associated components: free lipid, free receptor, and receptor bound to a protein from the bulk (Fig. 1a).

Schematics of the bulk-membrane coupled system.

a Schematic representation of scaffold proteins binding to the receptors anchored on a lipid bilayer. Each receptor (magenta) is tagged to a lipid (orange) in the supported lipid bilayers (SLBs). Scaffold proteins P (green) can bind to receptors R, forming a protein-receptor complex P R. We introduce two more effective components in the membrane: the lipids with a receptor without any protein bound but a water layer on top, and a lipid patch without a receptor but with the water layer on top. These components are termed “effective” because they also include water molecules that form a molecular layer adjacent to the membrane. For clarity, we do not depict these water molecules in the schematics, thereby providing a clearer view of the surface binding structure. We note that the height of the effective binding layer is denoted as h, which is also the height of protein-receptor complex units. In addition, hR represents the height of free receptors. Most importantly, the effective components are confined to the membrane and can only move diffusely parallel to it, or unbind and dissolve in the bulk. b shows the mathematical notations in our theoretical model. The volume fraction of proteins in the bulk is ϕ; ϕP R, ϕR are the area fraction of membrane-bound complexes and free receptors. Moreover, ϕ|z=0 is the volume fraction of protein at the membrane surface, which is denoted as S. So indicates the boundary surface of the bulk region V opposite to the membrane surface S. The boundary surfaces of the bulk region V , excluding S and So, are denoted by ∂V ; The membrane surface’s boundary is represented by ∂S.

The proteins in bulk can bind to receptors in the membrane, while membrane lipids are not accessible for protein binding. Note that the bulk does not contain any receptors. In our theory, we focus on the classical case of supported lipid bilayers (SLBs); however, we note that the theory can be straightforwardly extended to model real cell membranes even more realistically, e.g. by including more components and the effects of curvature. Since receptors are confined to the membrane, proteins that bind to these receptors can only diffuse within a layer adjacent to the membrane surface. This layer also contains solvent molecules, which are exchanged with the bulk when proteins bind. In the three-dimensional bulk V , a binary mixture of protein and solvent molecules is present. To capture processes occurring at the membrane surface, we introduce a quasi-two-dimensional domain S representing the membrane environment; see Fig. 1b for an illustration. Moreover, So indicates the boundary surface of bulk region V opposite to membrane surface S. The boundary surfaces of the bulk region V , excluding S and So, are denoted by ∂V . The membrane (one-dimensional) boundary is represented by ∂S.

To describe the setting outlined above, we propose a spatially coarse-grained model for bulk-membrane coupled systems. To this end, we introduce three effective components on a surface S: lipid with receptors bound to protein P R, lipid with receptors that are not bound to protein R, and lipid with only the water layer on top L. Proteins in the bulk are denoted by P . For simplicity, we consider an incompressible mixture where the molecular volume of solvent v and protein vP in bulk are constants. Similarly, the molecular areas of lipid vL, receptor vR and protein-receptor complex vP R are constants. The thermodynamics of such an incompressible mixture is characterized by the bulk volume fraction of the proteins ϕ(x, t), and the area fraction of the protein-receptor complex P R that is confined in the membrane ϕP R(x, t), and the area fraction, ϕR(x, t), of free receptor R on the membrane. Hence, under the incompressibility assumption, the solvent in the bulk occupies the fraction (1 − ϕ), and the free lipid in the membrane occupies the fraction (1 − ϕR − ϕP R). All volume fraction fields, in general, depend on position x = (x, y, z) and time t. The membrane is located at z = 0. We denote ϕ|z=0 as the volume fraction of protein at the boundary surface S = {(x, y, z = 0) ∈ V }. In contrast, the area fraction fields ϕP R(x, t) and ϕR(x, t) depend only on the in-plane coordinates x∥ = (x, y).

We consider the binding reaction of α proteins (labelled by P ) binding from the bulk to β receptors tagged to lipids in the membrane (labelled by R); see Fig. 1a for an illustration. This binding process leads to the complex P R and the release of (αn) solvent molecules (labelled by W ) to the bulk to conserve the volume of the molecules. The corresponding binding scheme is given as:

Here, α and β are the stoichiometric coefficients of the binding process. Volume conservation of the protein implies that the ratio of the molecular volumes between the protein and solvent is n = vP /vW . Note that the receptor R is an effective component made of the receptor tagged to lipids and the layer of solvent molecules on top. The height h of this layer is set by the complex , where is the complex molecular volume and vP R is the molecular area of the complex. For volume-conserved binding, the stoichiometric coefficients fulfill and β = vP R/vR.

The binding process is subject to mass conservation that can be expressed in terms of the total particle numbers Ni (i = R, P , P R): where the superscript 0 indicates the respective total particle numbers conserved in the binding process.

Both conservation laws can be rewritten in terms of the average area fractions for receptor and complex on the membrane, where A is the total area of membrane surface S, and the average protein volume fraction in the bulk : where is the conserved total receptor area fraction and ϕ0 is the conserved total protein volume fraction. The average area fractions are given as with i =R, P R and the average protein volume is .

Non-equilibrium thermodynamics with non-dilute surface binding

The area fraction fields ϕP R(x∥, t) and ϕR(x∥, t) and the bulk protein volume fraction field ϕ(x, t) obey conservation laws for membrane S and the bulk V : where j, jP R and jR are the diffusive fluxes of bulk protein P , complex P R and free receptor R. Moreover, r denotes the binding rate between bulk and membrane. We define ∇ = (∂x, ∂y, ∂z) as the gradient operator in the three-dimensional bulk and ∇|| = (∂x, ∂y) on the two-dimensional membrane surface.

Binding relates the protein flux n · j normal to the membrane S to the binding rate r: where α is the stoichiometric coefficient of the binding protein, and vP is the protein molecular volume and vP R is the molecular area of the bound complex. The fraction α vP /vP R equals the layer height h, providing the length scale related to binding from three to two dimensions. The remaining bulk boundaries are considered to be impermeable for all components with the diffusive flux vanishing: n · j = 0, x ∈ ∂V , So.

The total free energy F of the system is composed of different contributions: the free energy density fb(ϕ) in the bulk V , the free energy density fm(ϕP R, ϕR) at the membrane surface S, and coupling between bulk and membrane J (ϕ|z=0, ϕP R, ϕR). In specific, the total free energy reads:

The free energy penalties due to concentration gradients are characterized by κ, κP R, κR, and κP R-R in bulk and membrane, respectively.

In the following, we derive the kinetic equations that can relax toward thermodynamic equilibrium. In an isothermal system, the rate of change of the system entropy S(t) can be expressed in terms of the integrated change of the free energy densities fm, fb in membrane S and bulk V respectively, the coupling free energies J , as well as the reactive free energy fluxes jr1 and jr2 through the boundaries of bulk and membrane, respectively Zhao et al. (2024):

Using the conservation laws Eqs. (4)(a-c) and boundary conditions (4d), defining the reactive free energy fluxes jr1 = jμ/vP and jr2 = jP RμP R/vP R + jRμR/vR, we obtain for the entropy production rate, where we could define a pair of thermodynamic fluxes and forces. Using Onsager principle, we get the constitutive relation between fluxes and forces. Substituting the well-defined fluxes into the continuity equation (4)(a,b,c), we obtain the governing equations of the system as follows with boundary conditions:

We note that it is possible to have more dissipative fluxes on the boundaries S when the system evolves approaching the equilibrium state Zhao et al. (2024), e.g., (∂J /∂ϕ + n · κ∇ϕ)Λk = ∂tϕ|z=0, where Λk is the corresponding kinetic coefficient. Here, we only focus on the effects of binding rate r and thus set other fluxes to be zero for all times during the kinetics. The boundary condition ∂J /∂ϕ + n · κ∇ϕ = 0 in Eq. (8)a characterizes the interactions with the membrane surface and thereby also determine wetting behavior of bulk condensates. The second condition specifies that the opposite side of membrane So are neutral, i.e., the free energy is not affected by increasing or decreasing the bulk volume fractions adjacent to So. Moreover, we did not consider the cross-coupling interaction in the membrane.

We consider the standard mobility scalings, and consider for simplicity the mobility prefactors ΛP ,0, ΛP R,0, and ΛR,0 as constants.

The binding rate r obeys detailed balance of the rates Weber et al. (2019). However, by Onsager linear response thermodynamics Zhao et al. (2024), we can only obtain the binding rate to linear order in the thermodynamic force Δμ = μP R −β μR −αμ. To account for a general non-linear binding rate that also obeys detailed balance of the rates, we follow Ref. Van Kampen (1973) and write the binding rate where Λ is the binding rate coefficient. At thermodynamic equilibrium (Eqs. (14)), the binding rate r = 0.

Thermodynamic equilibrium with non-dilute surface binding

When each diffusive flux and the binding rate defined above vanish to zero, the system is at ther-modynamic equilibrium. At thermodynamic equilibrium, the total free energy F is minimal. There are constraints for the minimization since the total number of proteins and the total number of receptors at the membrane are conserved (Eqs. (2)). The minimum of the free energy fulfils where λ and λR denote Lagrange multipliers corresponding to the two respective conservation laws. Using Eq. (5), the variation of the total free energy is given as

We thus obtain the exchange chemical potentials of the proteins in the bulk μ, the protein-receptor complex μP R and the free receptor μR in the membrane:

Using Eqs. (13), we find that μP R = α λ+β λR, with λ = μ and λR = μR. As a result, binding equilibrium of the system composed of membrane and bulk is given as:

Here, the stoichiometric coefficients α and β reflect that effective component P R is composed of α proteins and β receptors.

Once the protein binds to the membrane-bound receptor and its local concentration exceeds a certain threshold, surface phase separation may occur, driven by the attractive interaction between protein–receptor complexes. When two phases, denoted as I and II, coexist in the membrane, phase equilibrium holds:

The total amount of receptors in the membrane is conserved (Eq. (3a)) and the conservation for two coexisting phases reads: where AI is the surface condensate area in the membrane, A = L2 is the total area of the membrane, and is the initial area fraction of the receptor in the membrane.

We use the boundary conditions at the membrane S, the opposite boundary surface So, and the rest of the boundaries enclosing the bulk, ∂V , at thermodynamic equilibrium in Eq. (8). Boundary conditions (8) and equilibrium conditions (14) are composite of non-dilute binding equilibrium conditions. In the following section, we will discuss the non-dilute surface binding in detail based on these equilibrium conditions.

Interplay between surface binding and surface phase separation

Non-dilute surface binding

In this section, we derive the general binding isotherm for equilibrium surface binding under non-dilute conditions. Unlike the classical Langmuir model Langmuir (1918), which assumes idealized dilute conditions and no intermolecular interactions, here we incorporate activity coefficients to account for deviations from ideality, such as interactions between different kinds of molecules. This approach leads to a generalized form of the binding isotherm that reduces to the classical Langmuir model under appropriate limiting assumptions.

To capture non-dilute behavior, we express the chemical potentials in terms of standard chemical potentials and activity coefficients: where the terms γ(ϕ), γR(ϕR, ϕP R), and γP R(ϕR, ϕP R) are activity coefficients capturing non-ideal effects, such as intermolecular interactions or internal energies. The constants μ0, μR,0, and μP R,0 are the corresponding standard chemical potentials that include the internal free energies of the components.

Substituting (15) into the binding equilibrium condition (14a), we rearrange to find, or equivalently, where

Equation (17a) represents a general non-dilute binding relation, where the effects of interactions at non-dilute conditions are accounted for by activity coefficients γ, γR and γP R. Moreover, Δμ0 denotes the differences of standard chemical potential.

To determine the surface coverage, we must also consider the conservation of the total amount of receptor (Eq. (3a)).

By combining this conservation law with the general binding relation (17a), we arrive at a generalized non-dilute binding isotherm. Upon choosing a specific form of the free energy densities fb, fm and J in Eq. (5), Eqs. (17) can only be solved numerically due to the non-linear dependence of the activity coefficients on all the area fractions of the components at non-dilute conditions.

For clarity, the non-dilute binding isotherm can be written in a simpler form for the special case of an unimolecular binding process (α = β = 1). Defining a quantity commonly studied in adsorption isotherms Ayawei et al. (2017), the binding fraction , we obtain denoting the generalized Langmuir coefficient. Here, Ω is the generalized binding affinity given in Eq. (17b), which describes the binding energy per molecule at non-dilute conditions.

In the ideal case where the binding sites are homogeneous and receptors are dilute, the effects of intermolecular interactions in the activity coefficients vanish in Eq. (18). As a result, one obtains the classical Langmuir isotherm Langmuir (1918), and KL reduces to the Langmuir constant KL = exp(Δμ0/kBT ), that is determined solely by the dilute binding affinity Δμ0.

In summary, starting from the general equilibrium conditions and incorporating conservation of mass and non-ideal interactions via activity coefficients, we have derived a general binding isotherm applicable to non-dilute conditions (17a). This result not only generalizes the theory of surface binding to non-dilute conditions recovering the classical Langmuir equation in the dilute limit, it also describes phase transitions at the binding surface leading to a heterogeneous distribution of bound molecules.

The non-dilute binding isotherm and surface phase separation

In this section, we investigate the intricate behaviors of non-dilute surface binding by varying bulk and surface compositions. Using Flory-Huggins free energy densities for the bulk-membrane system (Appendix 1), we assume unimolecular binding (α = β = 1) and examine the thermodynamic equilibrium satisfying the conditions of phase coexistence (14b)–(14d) and the generalized binding isotherm (18), we obtain with

To align with physical constraints and simplify computational implementation, we set the basic parameter values as follows: ωP R = −1, χ = 2.5, n = nP R = nR = 1, ω = ωR = 0, χP -L = 0, and χP R-R = 0.

Specifically, ωP R, the internal free energy coefficient, denotes the binding affinity of proteins to the membrane receptors. Setting ωP R = −1 reflects a moderate attractive interaction, consistent with observed protein-receptor binding in similar systems. The parameter χ = 2.5 ensures sufficient phase separation tendencies in the bulk, while n = nP R = nR = 1 assumes comparable molecular sizes between receptor-bound proteins, unbound receptors, and lipid molecules, simplifying the modeling framework. Finally, χP -L and χP R-R are set to zero to focus on dominant interactions and exclude secondary effects that do not significantly impact the observed behavior.

We first consider the scenario where phase separation does not occur on the surface, as the attractive interactions between bound proteins (χP R-L = 2) are insufficient to induce surface phase separation. Under these assumptions, our analysis reveals that the binding fraction θ increases monotonically with the bulk protein volume fraction ϕ in the non-dilute case (Fig. 2a) This behavior qualitatively resembles the dilute case, which corresponds to a scenario where the total receptor concentration is very low ; see conservation law (3a). However, in the non-dilute case, the binding fraction θ is significantly higher than in the dilute case as the total receptor concentration increases. This enhancement is a direct consequence of non-dilute binding. At very high bulk protein volume fractions (ϕ), the bulk undergoes phase separation, further altering the system’s behavior.

Non-dilute surface binding behaviors with varying surface and bulk compositions.

(a) Binding fraction curves are shown for three different initial receptor area fractions: (solid line), (dashed line) and (dotted line), under the assumption of no surface phase separation (χP R-L = 2). (b) With a stronger interaction strength (χP R-L = 3.2) and a fixed initial receptor area fraction , the homogeneous surface mixture (black line) undergoes phase separation into two distinct phases: a protein-rich phase (red line) and a protein-poor phase (blue line) as a function of the intermediate bulk volume fraction, ϕ. Additionally, when the bulk protein volume fraction (ϕ) exceeds the saturation level, a 3D bulk condensate forms (blue region) that coexist with 2D surface condensates on the membrane. (c) The phase diagram illustrates that intermediate receptor area fractions and large bulk protein volume fractions (ϕ) promote surface phase separation on the membrane. The basic parameter values are: ωP R = −1, χ = 2.5, n = nP R = nR = 1, ω = ωR = 0, χP -L = 0, and χP R-R = 0.

We now examine the case where the bound proteins at the surface can undergo surface phase separation, forming a protein-rich and a protein-poor phase leading to a heterogeneous distribution of bound proteins P R at the surface (see Fig. 2b). A sufficiently large bulk protein concentration is needed to ensure enough proteins are bound to receptors at the surface. Moreover, interactions need to be strong enough (e.g., χP R-L = 3.2), e.g., the bound proteins attract each other sufficiently strongly.

The effects of both the bulk protein volume fraction ϕ and the total amount of receptors on the surface are summarized as a phase diagram in Fig. 2c for a fixed value of χP R-L = 3.2. It shows that surface phase separation occurs in an intermediate range of both bulk protein and total receptor area fraction . A too-large value of bulk protein volume fraction triggers phase separation in the bulk independent of the receptor and surface binding. Moreover, a too-large amount of receptors on the surface leads to a homogeneous state for volume fraction fractions below bulk phase separation since the mixing entropy on the surface dominates. Similarly, if there are too few receptors on the surface, the enthalpic gains related to the interaction among bound proteins are also outcompeted by the mixing entropy. This overall indicates a reentrant behavior of surface phase separation when varying the total receptor area fraction .

Effects of binding affinity and interactions on surface phase separation

We also investigated how the binding affinity (ωP R) and the interaction strength between membrane-bound proteins (χP R-L) affect surface phase separation. We distinguish two scenarios (Fig. 3a): The three-dimensional bulk is homogeneous, or the bulk is phase-separated.

Surface behavior with varying binding affinity and membrane-bound interaction strength.

(a) Schematic illustrating bulk-membrane coupled system with surface condensates formed in the membrane S without and with bulk phase separation, respectively. (b) Phase diagram without bulk phase separation, indicating that strong binding affinity and (ωP R) interaction strength (χP R-L) among membrane-bound complexes promote surface phase separation. (c) If protein interaction in the bulk is strong and thus bulk phase separation occurs. With bulk condensates partially/completely wet on top of the membrane, the surface phase separation occurs in a much larger parameter regime. This illustrates that bulk protein properties, like interaction strength, also regulate the surface phase behavior. The basic parameter values we set in this study are: , (a) χ = 0, ϕ = csat = 0.145, (b) χ = 2.5, ϕ = csat = 0.145. Here, csat represents the bulk volume fraction at the saturation level.

In the absence of bulk phase separation, as shown in Fig. 3b, we find that upon increasing the binding affinity (ωP R) and membrane-bound interaction strength (χP R-L) can induce surface phase separation on the membrane, shifting the system from a homogeneous state to a phase-separated one. This result underscores the importance of binding affinity and membrane-bound complex interactions as a driving force for surface phase separation.

When bulk phase is phase-separated, as depicted in Fig. 3c, the parameter range of (ωP R) and (χP R-L) leading to surface separation broadens significantly. Under these conditions, surface phase separation and partial wetting are observed even at lower binding affinities (ωP R) compared to conditions without bulk phase separation. This finding suggests that interaction strength between proteins in the bulk, which gives rise to bulk phase separation, tunes surface phase separation. However, bulk phase separation is not a prerequisite for surface phase separation. Surface phase separation is dominated by the ability to recruit proteins from the bulk through binding that is in turn controlled by the binding affinity (ωP R) and the interaction strength between bound proteins (χP R-L).

Mechanism in tight junction formation

Tight junctions are important membrane-associated structures for epithelial and endothelial barrier formation. The phase separation of scaffold proteins ZO1 has been reported to be essential for tight junction assembly Beutel et al. (2019). However, how binding interplays with the spatiotemporal phase separation of ZO1 at the cell-cell contact site on the plasma membrane is not well understood. At the cell-cell interface, tight junction receptors including the Claudin family, the CTX family (JAMs) and the MARVEL family (Occludin, Tricellulin, MARVELD3), form multiple trans- and cis- interactions to mediate cell adhesion (Fig. 4a). Genetic mutation in these trans- and cis- interactions has been shown to induce tight junction and epithelial barriers deficiency Kostrewa et al. (2001); Zhao et al. (2018b); Heinemann and Schuetz (2019). We hypothesise that receptor oligomerization at the cell-cell interface drives ZO1 surface phase separation and tight junction formation. To test this idea, we use our theoretical model developed above to understand the mechanism for surface phase separation involved in the formation of tight junctions. We consider ZO1 as the scaffold proteins in the bulk, Claudin2 C terminal (CLDN2-ICD, referred as 1-mer here-after), which is one of the essential receptors for tight junction function, as the receptors anchored in the membrane. CDLN2-ICD were tagged with His tag and anchored to the membrane with DGS-NTA (Ni) lipid as a mimic of the membrane-associated receptors via His-Ni interaction (Fig. 4b). To mimic the receptor oligomerization happening at the cell-cell contact site, ySMF domain, which is known to form stable 14-mer Collins et al. (2003), was introduced to the CDLN2-ICD construct to get 14-meric receptors (referred as 14-mer hereafter). Scaffold proteins ZO1 were then added to the bulk. See Appendix 4 for methods and materials used in the in vitro experiments.

Illustration of tight junction initiation and in vitro reconstitution system.

(a) Schematic representation of tight junction initiation during cell-cell contact. Upon contact between Cell 1 and Cell 2, receptors (magenta) on the plasma membrane (PM) interact with scaffold proteins such as ZO1 (green) to form condensates that contribute to tight junction formation Beutel et al. (2019). (b) In vitro reconstitution system showing the binding dynamics of ZO1 proteins (green) with receptors (magenta) on a supported lipid bilayer. The His(10)-tagged receptors are anchored to DGS-NTA (Ni) lipid via His-Ni interaction and can bind and unbind with the ZO1 proteins in the bulk.

Now we use our theory for non-dilute binding to membrane-like surfaces and surface phase separation of bound proteins to model an in vitro reconstitution system for tight junction initiation. We will show that the theory will be instrumental to characterize binding affinities at non-dilute conditions and decipher the role of the oligomerization state of the receptor. As a crucial step in this analysis, we estimate the fundamental model parameter values for ZO1 protein binding to claudin receptors by integrating image analysis and experimental data (see Appendix 5 for details). These parameter values, which form the basis of subsequent modeling, are summarized in Table 1.

Receptor oligomerization supports surface phase transitions

We analyzed how the oligomerization state of the receptor affects the surface phase transition. In this study, we used two kinds of receptor oligomerization states: 1-mer and 14-mer. By fitting our model (14) to the experimental data (see Appendix 6) for both receptor states (Fig. 5c and 5d and Fig. 6c and 6d), we determined the changes in receptor binding affinity ωP R and self-interaction parameter χP R-L of membrane-bound ZO1 complexes. A key finding is that the binding affinity of the 14-meric receptor to ZO1 protein (ωP R = 3.83, Fig. 6) is much stronger compared to the monomeric receptor (ωP R = −1.09, Fig. 5). Furthermore, we found that the lower binding affinity of the 1-mer is dominantly caused by a reduced binding fraction, i.e., only a small fraction of 1-mer receptors can bind to ZO1 (see Appendix 7 for details); a mechanism previously reported in Ref. Erlendsson et al. (2019). Moreover, from the fits, we obtained that the bound ZO1-14-mer complexes strongly attract each other, i.e., they have a positive self-interaction parameter (Fig. 6). While the fitted self-interaction of the ZO1-1-mer (χP R-L = −0.4) complex was lower compared to the 14-mer(χP R-L = 1.09), a wide range of self-interaction values provide similar good fits (blue shade in Appendix 6-fig. 1c), indicating that in the low-affinity binding regime of the 1-mer receptor self-interactions at the surface play a minor role (see Appendix 6-fig. 1c and its related discussion).

Surface stays homogeneous with 1-mer receptor anchored on the membrane.

(a) Fluorescence images of ZO1 and 1-mer at 1-mer receptor titrations, ranging from 19386 to 102689 μm−2. (b) Fluorescence images of ZO1 and 1-mer showing the effect of ZO1 titration (0 to 800 nM) on 1-mer binding. Scalar bars in the bottom panels represent 2 μm. (c) The linear relationship between membrane-bound ZO1 density and receptor density on the membrane. This is because, in this study, the binding is saturated for all the receptor densities. In this figure, the dots with error bars represent the experimental data (quantified from figure (a)), and the solid line depicts the theoretical prediction. (d) Curve of membrane-bound ZO1 density as a function of ZO1 bulk density compared between experimental data (points) and theoretical prediction (line). The plateau is obtained from the binding fraction limit for 1-mer, c = 0.014. See Appendix 7 for details. The data is a fit to the membrane-bound ZO1 density observed in panel (b). (e) Phase diagram illustrating the homogeneous surface state across a physiological range of receptor and ZO1 densities.

Surface condensates form with 14-mer Receptor on the membrane

(a) Fluorescence images showing the distribution of 14-mer receptors and ZO1 across a range of receptor titrations, from 7 to 2744 μm−2. Surface phase separation occurs at intermediate receptor densities. Too few or too many receptors result in a homogeneous membrane phase: a dilute phase at low receptor densities and a dense phase at high receptor densities. (b) Fluorescence images of 14-mer receptors and ZO1 as a function of increasing ZO1 bulk concentration, from 4 to 14000 nM, demonstrating the emergence of phase-separated domains. Since receptor density remains constant during this titration, the area of surface condensates does not change with increasing ZO1 bulk concentration. Once the ZO1 bulk concentration exceeds the saturation level, bulk condensates form in the presence of surface condensates (refer to Fig. 3a for illustration). Scalar bars in the bottom panels represent 2 μm. (c) Membrane-bound ZO1 density as a function of receptor density on the membrane, showing how receptor density influences surface phase behavior. In this figure, the dots with error bars represent the experimental data (quantified from figure (a)), and the solid line depicts the theoretical prediction. (d) Membrane-bound ZO1 density dependence on ZO1 bulk concentration, comparing experimental data with theoretical predictions. The data is a fit to the membrane-bound ZO1 density observed in figure (b), revealing a transition from a homogeneous surface phase to a two-phase coexisting state (protein-poor and protein-rich phases). (e) Phase diagram with respect to the receptor density on the membrane and Bulk ZO1 density, indicating that 14-mer receptor promotes surface phase separation.

In summary, our results suggest that receptor oligomerization promotes surface phase separation through two synergistic mechanisms. First, increased binding affinity leads to a higher concentration of ZO1 on the membrane. Second, the self-interaction of the protein-receptor complex lowers the saturation concentration required for surface phase separation.

Using the parameters obtained from the fits, we determine the phase diagram for the monomeric (Fig. 5e) and 14-meric receptors (Fig. 6e). These phase diagrams show that there is no surface phase transition for monomeric receptors even for higher ZO1 concentrations, while there is a large region of surface phase separation for the 14-meric receptors.

Based on the agreements between experiments and our thermodynamic model, we conclude that a surface phase transition drives the formation of ZO1-rich surface condensates on membrane-like surfaces. The surface phase transition is controlled by the receptor’s oligomerization state, suggesting cell adhesion-induced oligomerization could act as the key switch for ZO1 surface condensation at cell-cell contacts.

Spatial pattern formation of ZO1 surface condensates

When proteins bound to the surface undergo phase separation, multiple condensates form and exhibit a slow ripening kinetics. Theoretical results on surface binding, surface condensates growth and ripening can be obtained from numerically solving the dynamic equations (22) in a three-dimensional bulk volume V with surface S (Figure 7(a), see Appendix 2 and Appendix 3 for details). Both theoretical predictions and reconstitution experiments reveal an initial stage where surface condensates form and grow rapidly, followed by a prolonged ripening phase characterized by slow kinetics (see Appendix 8, Fig. 1). During this ripening stage, the average condensate areas predicted by the theory align well with experimental observations. Importantly, over a time course of approximately 15 minutes, the amount of bound proteins remains nearly constant as the surface condensates ripen, with only a slight reduction in the number of condensates. To further validate the model, we compared the condensate patterns from experiments and theory at time t = 15 min for different receptor titration levels ranging from 25 to 3800 PN/μm2 and ZO1 titration levels from 4 to 100 nM (Fig. 7(b,c)). The patterns show quantitative agreement between theory and experiments, providing strong evidence that the theoretical framework effectively captures the non-dilute surface binding mechanisms and the condensate formation process on the surface. Furthermore, we observed that the receptor density on the membrane significantly influences the spatial patterns of surface condensates. At lower receptor densities, condensates tend to be sparse and less connected, while higher receptor densities promote larger and more interconnected condensates, consistent with predictions from the theoretical model (Fig. 7(b,c)).

Snapshots of ZO1 surface condensates from simulation (top) and experiments (bottom) when titrating the 14-mer receptor and ZO1 bulk concentration.

(a) Schematic of the system in our theoretical model, with 3D bulk volume V and membrane surface S, where surface condensates formed via surface phase separation. (b) Snapshots of ZO1 from simulation (top row) and experimental observations (bottom) under varying receptor titration levels, from 25 to 3800 μm−2, at 15 mins. (c) Snapshots of ZO1 from simulation (top row) and experimental observations under varying ZO1 concentration levels, from 4 to 100 nM, at 15 min. The experimental results match well with the theoretical predictions from perspectives of both condensate pattern and condensate size. Scale bars in the bottom panels represent 2 μm. we set the diffusivity coefficient value as Di = 0.01 μm2s−1, i = P , P R, R. The gradient coefficients κ = κP R = κR = 2 × 10−5μm−1.

Conclusion

In this study, we developed a comprehensive theoretical framework grounded in equilibrium ther-modynamics to describe how soluble scaffold proteins bind lipid membranes through interactions with membrane-embedded receptor proteins. The framework derives a generalized binding isotherm, based on the thermodynamic theory for non-dilute surface binding. This isotherm extends classical adsorption models by accounting for the non-dilute nature of membrane-bound complexes and their interactions. Additionally, a kinetic model was derived from non-equilibrium thermodynamics to investigate spatiotemporal patterns of surface condensates.

Using this framework, we explored the behaviors of non-dilute surface binding under varying bulk and surface compositions, binding affinities, and interaction strengths between membrane-bound complexes. The model also enabled us to simulate and analyze surface phase transitions in an in vitro reconstitution system for tight junction initiation. A critical insight from this work is the role of receptor oligomerization in facilitating surface phase transitions, highlighting its importance in membrane-associated condensation phenomena.

Our findings demonstrate that receptor density significantly influences the spatial patterns of surface condensates. At lower receptor densities, condensates are sparse and less connected, whereas higher receptor densities promote larger and more interconnected structures. Moreover, the patterns observed in simulations show quantitative agreement with experimental results, providing strong evidence that the theoretical framework accurately captures the mechanisms of non-dilute surface binding and condensate formation processes.

In summary, this work bridges theoretical modeling and experimental validation, advancing our understanding of non-dilute surface binding and phase separation on biological membranes. This framework offers a versatile tool for further investigating multi-component systems, membrane curvature effects, and the dynamic organization of cellular interfaces, with potential applications in both biological research and biomimetic design.

eLife Assessment

This important study presents a compelling theoretical framework for understanding phase separation of membrane-bound proteins, with a focus on the organization of tight junction components. By incorporating non-dilute binding effects into thermodynamic models and validating the model's predictions with in vitro experiments on the tight junction protein ZO-1, the authors provide a quantitative tool that will be of interest for biologists interested in membrane-associated condensates. While further clarification of model assumptions and broader mechanistic context would strengthen the work even further, the combination of theory and experiment here is robust and a key advancement in the field.

Reviewer #1 (Public review):

Summary:

Biomolecular condensates are an essential part of cellular homeostatic regulation. In this manuscript, the authors develop a theoretical framework for the phase separation of membrane-bound proteins. They show the effect of non-dilute surface binding and phase separation on tight junction protein organization.

Strengths:

It is an important study, considering that the phase separation of membrane-bound molecules is taking the center stage of signaling, spanning from immune signaling to cell-cell adhesion. A theoretical framework will help biologists to quantitatively interpret their findings.

Weaknesses:

Understandably, the authors used one system to test their theory (ZO-1). However, to establish a theoretical framework, this is sufficient.

Reviewer #2 (Public review):

Summary:

The authors present a clear expansion of biophysical (thermodynamic) theory regarding the binding of proteins to membrane-bound receptors, accounting for higher local concentration effects of the protein. To partially test the expanded theory, the authors perform in vitro experiments on the binding of ZO1 proteins to Claudin2 C-terminal receptors anchored to a supported lipid bilayer, and capture the effects that surface phase separation of ZO1 has on its adsorption to the membrane.

Strengths:

(1) The derived theoretical framework is consistent and largely well-explained.

(2) The experimental and numerical methodologies are transparent.

(3) The comparison between the best parameterized non-dilute theory is in reasonable agreement with experiments.

Weaknesses:

(1) In the theoretical section, what has previously been known, compared to which equations are new, should be made more clear.

(2) Some assumptions in the model are made purely for convenience and without sufficient accompanying physical justification. E.g., the authors should justify, on physical grounds, why binding rate effects are/could be larger than the other fluxes.

(3) I feel that further mechanistic explanation as to why bulk phase separation widens the regime of surface phase separation is warranted.

(4) The major advantage of the non-dilute theory as compared with a best parameterized dilute (or homogenous) theory requires further clarification/evidence with respect to capturing the experimental data.

(5) Discrete (particle-based) molecular modelling could help to delineate the quantitative improvements that the non-dilute theory has over the previous state-of-the-art. Also, this could help test theoretical statements regarding the roles of bulk-phase separation, which were not explored experimentally.

(6) Discussion of the caveats and limitations of the theory and modelling is missing from the text.

Author response:

The following is the authors’ response to the original reviews.

Reviewer #1 (Public review):

Summary:

Biomolecular condensates are an essential part of cellular homeostatic regulation. In this manuscript, the authors develop a theoretical framework for the phase separation of membrane-bound proteins. They show the effect of non-dilute surface binding and phase separation on tight junction protein organization.

Strengths:

It is an important study, considering that the phase separation of membrane-bound molecules is taking the center stage of signaling, spanning from immune signaling to cell-cell adhesion. A theoretical framework will help biologists to quantitatively interpret their findings.

Weaknesses:

Understandably, the authors used one system to test their theory (ZO-1). However, to establish a theoretical framework, this is sufficient.

We acknowledge this limitation. While we agree that additional systems would strengthen the generality of our theory, we note that the focus of this work is to introduce and validate a theoretical framework. As the reviewer notes, this is sufficient for establishing the framework. Nonetheless, we are open to further collaborations or future studies to test the model with other systems.

Reviewer #2 (Public review):

Summary:

The authors present a clear expansion of biophysical (thermodynamic) theory regarding the binding of proteins to membrane-bound receptors, accounting for higher local concentration effects of the protein. To partially test the expanded theory, the authors perform in vitro experiments on the binding of ZO1 proteins to Claudin2 C-terminal receptors anchored to a supported lipid bilayer, and capture the effects that surface phase separation of ZO1 has on its adsorption to the membrane.

Strengths:

(1) The derived theoretical framework is consistent and largely well-explained.

(2) The experimental and numerical methodologies are transparent.

(3) The comparison between the best parameterized non-dilute theory is in reasonable agreement with experiments.

Weaknesses:

(1) In the theoretical section, what has previously been known, compared to which equations are new, should be made more clear.

We have revised the theory section to clearly distinguish previously established formulations from novel contributions following equation (4), which is .

(2) Some assumptions in the model are made purely for convenience and without sufficient accompanying physical justification. E.g., the authors should justify, on physical grounds, why binding rate effects are/could be larger than the other fluxes.

For our problem, binding is relevant together with diffusive transport in each phase. Each process is accompanied by kinetic coefficients that we estimate for the experimental system. For the considered biological systems (and related ones), it is difficult to determine whether other fluxes (see, e.g., Eq. 8(e)) have relaxed or not. We note that their effects are, of course, included in the kinetic model applied to the coarsening of ZO1 surface condensates as boundary conditions. But we cannot exclude that the corresponding kinetic coefficient in the actual biological system is large enough such that, e.g., Eq. (9e) does not vanish to zero “quasi-statically”. We have now added a sentence to the outlook highlighting the relevance of testing those flux-force relationships in biological systems.

(3) I feel that further mechanistic explanation as to why bulk phase separation widens the regime of surface phase separation is warranted.

We have discussed the mechanistic explanation related to bulk protein interaction strength in the manuscript in the section: “Effects of binding affinity and interactions on surface phase separation”. We explained how the bulk interaction parameter affects the binding equilibrium.

(4) The major advantage of the non-dilute theory as compared with a best parameterized dilute (or homogenous) theory requires further clarification/evidence with respect to capturing the experimental data.

We thank reviewer for this helpful question. To address this point, we have added new paragraphs in the conclusion section, which explicitly discuss the necessity of employing the non-dilute theory for interpreting the experimental data.

(5) Discrete (particle-based) molecular modelling could help to delineate the quantitative improvements that the non-dilute theory has over the previous state-of-the-art. Also, this could help test theoretical statements regarding the roles of bulk-phase separation, which were not explored experimentally.

We appreciate the suggestion and agree that such modeling would be valuable. However, this is beyond the scope of the current study.

(6) Discussion of the caveats and limitations of the theory and modelling is missing from the text.

We sincerely appreciate the reviewer’s helpful comment. We have added a discussion in the conclusion section outlining the caveats and limitations of our modeling approach.

Reviewing Editor Comments:

Upon discussing with the reviewers, we feel that this manuscript could significantly be improved if testing the model with a different model system (beyond ZO1/tight junctions), in which case we foresee that we could enhance the strength of evidence from "compelling" to "exceptional". But of course, this is up to the authors to go for it or not, the paper is already very good.

Reviewer #2 (Recommendations for the authors):

(1) Lines 132-134: Re-word, the use of "complex" is confusing.

We have rephrased the sentence for clarity. The revised version reads: ṽ𝑃𝑅 are the molecular volume and area of the protein-receptor complex ѵ𝑃𝑅, respectively”, and the changes have been in the revised manuscript.

(2) Line 154 use of ""\nu"" for volume and area could be avoided for better clarity.

We thank the reviewer for this helpful suggestion. We have removed the statement involving ""\nu"" as these quantities have already been defined in the preceding context.

(3) Line 158 the total "Helmholtz" free energy F...

We have added the word "Helmholtz" to the sentence.

(4) Line 160 typo "In specific,..."

We carefully checked this sentence but could not identify a typo.

(5) For equation 5 explain the physical origins of each term, or provide a reference if this equation is explained elsewhere.

Thank you very much for your valuable suggestions. We have carefully rephrased Equation (5) and added a paragraph immediately afterward to provide a detailed explanation of its physical meaning.

(6) Derivation on lines 163-174 is poorly written. Make the logical flow between the equations clearer.

We greatly appreciate your insightful suggestions. Equation (6) has been carefully revised for clarity, and the explanation has been rewritten to ensure better readability. All modifications are Done.

(7) Define bold "t" in Equation 6.

The variable “t” has been explicitly defined in the context for clarity.

(8) In equations. 7b-7c the nablas (gradients) should be the 2D versions.

We have updated the gradient operators in Equations (7b) and (7c) [Eq. (9) in revised manuscript] to their 2D forms for consistency.

(9) Line 190, avoid referring to the future Equation 14, and state in words what is meant by "thermodynamic equilibrium".

We have added the explanation of “thermodynamic equilibrium” and remove the reference to equation accordingly.

(10) In Equation 11 you don't explain what you are doing ( which is a perturbation around the minimum of the free energy).

We have revised the paragraph before equation (11) [Eq. (13) in revised manuscript] to clarify that the expression represents a perturbation around the minimum of the free energy.

(11) In Equation 12, doesn't this also depend on how you have written equation 6 (not just equation 5).

Eq. (12) [Eq. (14) in revised manuscript] is derived directly from the variation of the total free energy F. In contrast, Eq. (6) contains the time derivative of free energies that were not written in their final form. In the revised version, we have now given the conjugate forces and fluxes in Eqs. (7) and (8) for clarity.

(12) Line 206 specify the threshold of local concentration (or provide a reference).

We have specified the threshold of local concentration in the revised text, and the corresponding statement has been highlighted.

(13) Line 223 is the deviation from ideality captured in a pair-wise fashion? I presume it does not account for N many-body interactions?

Yes, our model is formulated within a mean-field framework that incorporates pairwise (second order) interaction coefficients. For example, 𝜒𝑃𝑅 -𝑅 characterizes the interaction between the complex 𝑃𝑅 and the free receptor 𝑅, 𝜒𝑅 -L the interaction between free receptor 𝑅 and free lipid 𝐿, 𝜒𝑃𝑅-𝐿 the interaction between complex 𝑃𝑅and free lipid 𝐿. We have stressed this choice of free energy in the revised manuscript.

(14) Line 274, how do the authors know the secondary effects (of which they should mention a few) do not significantly impact the observed behaviour?

We sincerely thank the reviewer for the helpful comment. First, the parameters 𝜒𝑅 -L and 𝜒𝑃𝑅 -𝑅 are not essential based on the experimental observations. For more information, please see our revised paragraph on the choice of the specific parameter values, which has been in the following Eq. (21).

(15) It's not clear how Figures 3 b and c are generated with reference to which parameters are changed to investigate with/without bulk phase separation.

To improve clarity, we have revised Figure 3 to display the corresponding parameter values directly in each panel. Figures 3b and 3c were generated by computing the surface binding curves (as shown in Fig. 2) for each binding affinity 𝜔𝑃𝑅 and membrane-complex interaction strength 𝜒𝑃𝑅-𝐿, under different bulk interaction strengths chi, to compare the cases with and without bulk phase separation.

(16) The jump between theory and the "Mechanism in ..." section is too much. The authors should include the biological context of tight junctions and ZO1 in the main introduction.

We appreciate the reviewer’s suggestion. Following this comment, we have added an extended discussion in the main introduction to provide the necessary biological context of tight junctions and ZO1. In addition, we inserted new bridging paragraphs between the theoretical section and the section “Mechanism in tight junction formation” to create a smoother transition from theory to experiments. These revisions help to better connect the theoretical framework with the biological phenomena discussed in the later section.

Version 2

Abstract

Surface binding and surface phase separation of cytosolic scaffold proteins on lipid membranes are involved in many cellular processes, such as cell signaling, cell adhesion, and cortex regulation. However, the interplay between surface binding and surface phase separation is poorly understood. In this work, we study this interplay by deriving a general thermodynamic model and applying it to in vitro reconstitution experiments of membrane-binding proteins involved in tight junction initiation. Our theory extends the classical surface binding isotherm to account for non-dilute and heterogeneous conditions where components can phase separate. We use our theory to demonstrate how surface phase separation is governed by the interaction strength among membrane-bound scaffold proteins and their binding affinity to the membrane surface. Comparing the theory to reconstitution experiments, we show that tuning the oligomerization state of the adhesion receptors in the membrane controls surface phase transition and patterning of the scaffold protein ZO1. These findings suggest a fundamental role of the interplay between non-dilute surface binding and surface phase separation in forming the tight junction. More broadly, our work highlights non-dilute surface binding and surface phase separation as a common organizational principle for membrane-associated structures in living cells.

Introduction

Liquid-liquid phase separation (LLPS) has emerged as a fundamental principle governing the spatial organization of cellular components into dynamic, membrane-less compartments termed biomolecular condensates Brangwynne et al. (2011); Mahen and Venkitaraman (2012); Gall (2003); Brang-wynne et al. (2009); Fritsch et al. (2021); JR and R. (2009); CJ and R. (2012). Such condensates create distinct microenvironments that influence a wide range of cellular functions, including signaling Su et al. (2016); Wang et al. (2021); Xiao et al. (2022), transcription Boija et al. (2018); Lu et al. (2020), and stress responses Riback et al. (2017).

While LLPS has been extensively studied in the cytoplasm, protein phase separation is also involved in many biological processes happening at membrane surfaces such as signaling Li et al. (2012); CASE et al. (2019); Zeng et al. (2019); Li et al. (2022); Zeng et al. (2018); Su et al. (2016); Zheng et al. (2022), synapse formation Zeng et al. (2018), cell adhesion Beutel et al. (2019); Pombo-García et al. (2024); Wang et al. (2021) and cortex regulation CASE et al. (2019); Wiegand and Hyman (2020). Protein condensation at membranes is often induced by binding of multivalent scaffold proteins present in the cytoplasm to integral membrane proteins or lipids on the surface of the membrane. Membrane binding of scaffold proteins can result in surface condensation far below the concentration required for phase separation in the bulk, providing spatial-temporal control over the condensation process. For instance, tight junctions are essential for maintaining epithelial barrier integrity by forming continuous adhesive belts around cell perimeters. Recent studies have revealed that the condensation of ZO1 scaffold proteins is a critical step in tight junction assembly Beutel et al. (2019). Specifically, ZO1 initially binds to the receptors in the membrane and undergoes surface condensation at sites of cell-cell contact, a process initiated by the oligomerization of adhesion receptors. These findings underscore the interplay between membrane binding and surface phase separation Sun et al. (2024).

How membrane binding and surface phase separation influence each other is not well understood. Classical surface binding theories, such as the Langmuir isotherm Langmuir (1918), assume dilute concentrations where binding sites are independent and non-interacting. These assumptions are not valid for phase separating systems, where surface densities are non-dilute and intermolecular interactions among membrane-bound molecules become important. More recently, non-dilute binding and phase separation have been incorporated into theoretical models Zhao et al. (2021, 2024); Liese et al. (2025). However, these models still rely on simplifying assumptions such as uniform binding sites, which is not matching the reality of biological membranes.

Binding of proteins to biological membranes is typically tightly controlled in space and time by either increasing local densities of binding sites such as in the formation of cell adhesion complexes Otani et al. (2019); Zhao et al. (2018b), or by increasing binding affinities for example by phosphorylation Li et al. (2012). To accurately describe such phenomena, theoretical models must incorporate both non-homogeneous and non-dilute surface densities of proteins at the membrane. While a recent thermodynamic model has linked binding heterogeneity to prewetting phenomena Rouches et al. (2021), developing a theory that integrates these heterogeneous, non-dilute conditions and captures the potential for surface-bound phase transitions remains crucial for a comprehensive understanding of membrane-associated assemblies.

In this work, we present a theoretical framework for the interplay between surface binding and surface phase separation that includes heterogeneous and non-dilute membrane surface densities. We show how the classical binding isotherm must be fundamentally modified when molecules bound to membrane receptors interact strongly and undergo surface condensation. We derived a generalized binding isotherm which extends classical adsorption models by accounting for the non-dilute nature of surface-bound complexes and their interactions. Using this framework, we explored the behaviors of non-dilute surface binding under varying bulk and surface compositions, binding affinities, and interaction strengths between membrane bound complexes. To illustrate the applicability of our theory, we focus on in vitro reconstituted membrane systems that model components of tight junctions. We explain how oligomerization states influence membrane organization. More generally, our theory reveals how non-dilute surface binding and membrane-associated phase separation interact to influence membrane organization, shedding light on a fundamental physicochemical mechanism through which cells regulate membrane associated assembly processes.

Model for protein binding to receptors on a membrane

In the first part of this work, we discuss a theoretical framework that describes binding of soluble scaffold proteins to lipid membranes via interactions with receptor proteins embedded in a membrane. In addition, the scaffold proteins can undergo (surface) phase separation, which results in formation of protein-dense and protein-dilute regions on the membrane (Fig. 1a,b). The theoretical framework builds upon previous works on the interplay of surface-associated phase transition (surface phase separation, prewetting) and homogeneous surface binding Zhao et al. (2021, 2024). However, previous models do not involve a receptor component, reducing the stable coexistence of domains poor and rich in bound protein to a specific value of the bulk protein concentration where the phase transition occurs. However, this behavior is not consistent with in vitro experiments where phase coexistence exists over a large range of bulk protein concentrations. In this light, we propose an extended but still minimal model in terms of the number of membrane-associated components: free lipid, free receptor, and receptor bound to a protein from the bulk (Fig. 1a).

Schematics of the bulk-membrane coupled system.

a Schematic representation of scaffold proteins binding to the receptors anchored on a lipid bilayer. Each receptor (magenta) is tagged to a lipid (orange) in the supported lipid bilayers (SLBs). Scaffold proteins P (green) can bind to receptors R, forming a protein-receptor complex PR. We introduce two more effective components in the membrane: the lipids with a receptor without any protein bound but a water layer on top, and a lipid patch without a receptor but with the water layer on top. These components are termed “effective” because they also include water molecules that form a molecular layer adjacent to the membrane. For clarity, we do not depict these water molecules in the schematics, thereby providing a clearer view of the surface binding structure. We note that the height of the effective binding layer is denoted as h, which is also the height of protein-receptor complex units. Most importantly, the effective components are confined to the membrane and can only move diffusely parallel to it, or unbind and dissolve in the bulk. b shows the mathematical notations in our theoretical model. The volume fraction of proteins in the bulk is ϕ; ϕPR, ϕR are the area fraction of membrane-bound complexes and free receptors. Moreover, ϕ| z=0 is the volume fraction of protein at the membrane surface, which is denoted as S. So indicates the boundary surface of the bulk region V opposite to the membrane surface S. The boundary surfaces of the bulk region V, excluding S and So, are denoted by ∂V ; The membrane surface’s boundary is represented by ∂S.

The proteins in bulk can bind to receptors in the membrane, while membrane lipids are not accessible for protein binding. Note that the bulk does not contain any receptors. In our theory, we focus on the classical case of supported lipid bilayers (SLBs); however, we note that the theory can be straightforwardly extended to model real cell membranes even more realistically, e.g. by including more components and the effects of curvature. Since receptors are confined to the membrane, proteins that bind to these receptors can only diffuse within a layer adjacent to the membrane surface. This layer also contains solvent molecules, which are exchanged with the bulk when proteins bind. In the three-dimensional bulk V, a binary mixture of protein and solvent molecules is present. To capture processes occurring at the membrane surface, we introduce a quasi-two-dimensional domain S representing the membrane environment; see Fig. 1b for an illustration. Moreover, So indicates the boundary surface of bulk region V opposite to membrane surface S. The boundary surfaces of the bulk region V, excluding S and So, are denoted by ∂V. The membrane (one-dimensional) boundary is represented by ∂S.

To describe the setting outlined above, we propose a spatially coarse-grained model for bulk-membrane coupled systems. To this end, we introduce three effective components on a surface S: lipid with receptors bound to protein PR, lipid with receptors that are not bound to protein R, and lipid with only the water layer on top L. Proteins in the bulk are denoted by P, and solvent denoted by W. For simplicity, we consider an incompressible mixture where the molecular volumes of solvent v and protein vP in bulk and protein-receptor complex in the membrane are constants. Similarly, the molecular areas of lipid vL, receptor vR and protein-receptor complex vPR are constants. The thermodynamics of such an incompressible mixture is characterized by the bulk volume fraction of the proteins ϕ(x, t), and the area fraction of the protein-receptor complex PR that is confined in the membrane ϕPR(x∥, t), and the area fraction, ϕR(x∥, t), of free receptor R on the membrane. Hence, under the incompressibility assumption, the solvent in the bulk occupies the fraction (1 − ϕ), and the free lipid in the membrane occupies the fraction (1 − ϕR − ϕPR). All volume fraction fields, in general, depend on position x = (x, y, z) and time t. The membrane is located at z = 0. We denote ϕ| z=0 as the volume fraction of protein at the boundary surface S = {(x, y, z = 0) ∈ V }. In contrast, the area fraction fields ϕPR(x∥, t) and ϕR(x∥, t) depend only on the in-plane coordinates x∥ = (x, y).

We consider the binding reaction of α proteins (labelled by P) binding from the bulk to β receptors tagged to lipids in the membrane (labelled by R); see Fig. 1a for an illustration. This binding process leads to the complex PR and the release of (αn) solvent molecules (labelled by W) to the bulk to conserve the volume of the molecules. The corresponding binding scheme is given as: Here, α and β are the stoichiometric coefficients of the binding process. Volume conservation of the protein implies that the ratio of the molecular volumes between the protein and solvent is n = vP /vW. Note that the receptor R is an effective component made of the receptor tagged to lipids and the layer of solvent molecules on top. The height h of this layer is set by the complex , where are the molecular volume and area of the protein-receptor complex PR, respectively. For volume-conserved binding, the stoichiometric coefficients fulfill and β = vPR/vR.

The binding process is subject to mass conservation that can be expressed in terms of the total particle numbers Ni (i = R, P, PR): where the superscript 0 indicates the respective total particle numbers conserved in the binding process.

Both conservation laws can be rewritten in terms of the average area fractions for receptor and complex on the membrane, where A is the total area of membrane surface S, and the average protein volume fraction in the bulk : where is the conserved total receptor area fraction and is the conserved total protein volume fraction. The average area fractions are given as with i = R, P R and the average protein volume is .

Non-equilibrium thermodynamics with non-dilute surface binding

The area fraction fields ϕPR(x∥, t) and ϕR(x∥, t) and the bulk protein volume fraction field ϕ(x, t) obey conservation laws for membrane S and the bulk V : where j, jPR and jR are the diffusive fluxes of bulk protein P, complex PR and free receptor R. Moreover, r denotes the binding rate between bulk and membrane. It is worth noting that ϕR(x∥, t) denotes the area fraction of free receptors in the membrane. Accounting for the free receptor is crucial for the later comparison to experimental data. Moreover, this additional field is also a key extension compared to previous theoretical model on surface phase separation Zhao et al. (2021, 2024). The temporal evolution of ϕR is governed by Eq. (4c). We define ∇ = (∂x, ∂y, ∂z) as the gra-dient operator in the three-dimensional bulk and ∇ ||= (∂x, ∂ y) on the two-dimensional membrane surface.

Binding relates the protein flux n · j normal to the membrane S to the binding rate r: where α is the stoichiometric coefficient of the binding protein. The fraction α vP /vPR equals the layer height h, providing the length scale related to binding from three to two dimensions. The remaining bulk boundaries are considered to be impermeable for all components with the diffusive flux vanishing: n · j = 0, x ∈ ∂V, So.

The total Helmholtz free energy F of the system is composed of different contributions: the free energy in the bulk V, the free energy at the membrane surface S, and coupling free energy between bulk and membrane. In specific, the total free energy reads: where fb(ϕ) is the bulk free energy density in the bulk volume V, fm(ϕPR, ϕR) denotes the bulk free energy density at the membrane surface S, and J (ϕ | z=0, ϕPR, ϕR) is the coupling free energy density between bulk and membrane. In addition, we consider the interfacial free energies introduced by the concentration gradients at the phase interfaces, which are characterized by κ, and κR, κPR, κPR-R, in the bulk and the membrane, respectively.

In the following, we derive the kinetic equations that can relax toward thermodynamic equilibrium. In an isothermal system, the rate of change of the system entropy S(t) is negatively propositional to the change rate of the total Helmholtz free energy F Zhao et al. (2024): where jr1 and jr2 denote the non-dissipative, commonly referred to as reactive free energy fluxes. They describe free energy fluxes through the boundaries of bulk and membrane, respectively. Here, t denotes the normal vector to the boundary ∂S in the membrane.

Using the conservation laws Eqs. (4), boundary conditions (4d), and defining the reactive free energy fluxes jr1 = jμ/vP and jr2 = jPRμPR/vPR + jRμR/vR, we obtain the entropy production rate as a product of a pair of thermodynamic fluxes and forces Zhao et al. (2024) as follows: Then, we identify the conjugate thermodynamic fluxes and forces as products in the brackets above.

Consistent with Onsager’s principle and the second law of thermodynamics, we define the constitutive relation between these thermodynamic fluxes and forces to linear order as follows: ensuring that the entropy of the system will increase when the system approaches thermodynamic equilibrium. Here, Λ denotes the mobility of binding and ΛP, ΛPR and ΛR are the diffusive mobility coefficients in the bulk and the membrane, respectively. Moreover, and are Onsager coefficients.

Substituting the well-defined fluxes into the conservation laws (4)(a,b,c), we obtain the governing equations of the system as follows: with boundary conditions: Here, we assume that there are no dissipative fluxes on the boundary S when the system approaches the equilibrium state. i.e. we set and . We note that it is possible to have dissipative fluxes Zhao et al. (2024), e.g., −(∂J/∂ϕ| z=0 + n · κ∇ϕ)Λk = ∂tϕ | z=0, where Λk is the corresponding kinetic coefficient. Here, we focus on the effect of the binding rate r leading to compositional changes on a time-scale τ, while assuming that other fluxes relax rapidly and remain effectively at equilibrium during the entire kinetic process. Specifically, when Λκτ ≫ 1, ∂J /∂ϕ|z=0 + n · κ∇ϕ ≃ 0. To our knowledge, Λκ has not been determined for biological systems.

However, the relaxation is expected to be fast compared to binding processes since the normal concentration gradients n · ∇ϕ are typically confined to microscopic length scales on which diffusion enables quick relaxation.

The boundary condition ∂J /∂ϕ| z=0+n·κ∇ϕ = 0 in Eq. (10a) characterizes the interactions with the membrane surface and thereby also determine wetting behavior of bulk condensates. The second condition specifies that the opposite side of the membrane So are neutral, i.e., the free energy is not affected by increasing or decreasing the bulk volume fractions adjacent to So. Moreover, we did not consider the cross-coupling interaction in the membrane.

We consider the standard mobility scalings, and consider for simplicity the mobility prefactors ΛP,0, ΛPR,0, and ΛR,0 as constants.

The binding rate r obeys detailed balance of the rates Weber et al. (2019). However, by Onsager linear response thermodynamics Zhao et al. (2024), we can only obtain the binding rate to linear order in the thermodynamic force Δμ = μPR −β μR −αμ. To account for a general non-linear binding rate that also obeys detailed balance of the rates, we follow Ref. Van Kampen (1973) and write the binding rate where Λ is the binding rate coefficient. At thermodynamic equilibrium, there are no net flows of matter or energy within the system; that is, the binding flux satisfies r = 0.

Thermodynamic equilibrium with non-dilute surface binding

At thermodynamic equilibrium, the total Helmholtz free energy functional F (ϕ, ϕP, ϕR, ϕPR) is minimal. Mathematically, this corresponds to δF = 0. In other words, the free energy does not changes in response to small changes of the fields ϕ, ϕP, ϕR, ϕPR and thus the linear order perturbations δϕ, δϕP, δϕR, δϕPR, of the free energy vanish to zero. The system satisfies two conservation laws: the total number of proteins and the total number of receptors at the membrane are conserved (Eqs. (2)). Overall, at thermodynamic equilibrium, the first variation of the total Helmholtz free energy F, subject to these conservation constraints, vanishes: where λ and λR denote Lagrange multipliers corresponding to the two respective conservation laws. Using Eq. (5), the variation of the total Helmholtz free energy is given as We thus obtain the exchange chemical potentials of the proteins in the bulk μ, the protein-receptor complex μPR and the free receptor μR in the membrane: Using Eqs. (15), we find that μPR = α λ+β λR, with λ = μ and λR = μR. As a result, binding equilibrium of the system composed of membrane and bulk is given as: Here, the stoichiometric coefficients α and β reflect that effective component PR is composed of α proteins and β receptors.

Once the protein binds to the membrane-bound receptor and its local concentration exceeds a certain threshold , surface phase separation occurs, driven by the attractive interaction between protein–receptor complexes. When two phases, denoted as I and II, coexist in the membrane, phase equilibrium holds: We note that the threshold value for surface phase separation identified above coincides with the volume fraction of the Protein–Receptor complex, i.e., . The total amount of receptors in the membrane is conserved (Eq. (3a)) and the conservation for two coexisting phases reads: where AI is the surface condensate area in the membrane, A = L2 is the total area of the membrane, and is the initial area fraction of the receptor in the membrane.

We use the boundary conditions at the membrane S, the opposite boundary surface So, and the rest of the boundaries enclosing the bulk, ∂V, at thermodynamic equilibrium in Eq. (10). Boundary conditions (10) and equilibrium conditions (16) are composite of non-dilute binding equilibrium conditions. In the following section, we will discuss the non-dilute surface binding in detail based on these equilibrium conditions.

Interplay between surface binding and surface phase separation

Non-dilute surface binding

In this section, we derive the general binding isotherm for equilibrium surface binding under non-dilute conditions. Unlike the classical Langmuir model Langmuir (1918), which assumes idealized dilute conditions and no intermolecular interactions, here we incorporate activity coefficients to account for deviations from ideality, such as interactions between different kinds of molecules. This approach leads to a generalized form of the binding isotherm that reduces to the classical Langmuir model under appropriate limiting assumptions.

To capture non-dilute behavior, we express the chemical potentials in terms of standard chemical potentials and activity coefficients: where the terms γ(ϕ), γR(ϕR, ϕPR), and γPR(ϕR, ϕPR) are activity coefficients capturing non-ideal effects, such as intermolecular interactions or internal energies. The constants μ0, μR,0, and μPR,0 are the corresponding standard chemical potentials that include the internal free energies of the components.

Substituting (17) into the binding equilibrium condition (16a), we rearrange to find, or equivalently, where Equation (19a) represents a general non-dilute binding relation, where the effects of interactions at non-dilute conditions are accounted for by activity coefficients γ, γR and γPR. Moreover, Δμ0 denotes the differences of standard chemical potential.

To determine the surface coverage, we must also consider the conservation of the total amount of receptor (Eq. (3a)).

By combining this conservation law with the general binding relation (19a), we arrive at a generalized non-dilute binding isotherm. Upon choosing a specific form of the free energy densities fb, fm and J in Eq. (5), Eqs. (19) can only be solved numerically due to the non-linear dependence of the activity coefficients on all the area fractions of the components at non-dilute conditions.

For clarity, the non-dilute binding isotherm can be written in a simpler form for the special case of an unimolecular binding process (α = β = 1). Defining a quantity commonly studied in adsorption isotherms Ayawei et al. (2017), the binding fraction , we obtain denoting the generalized Langmuir coefficient. Here, Ω is the generalized binding affinity given in Eq. (19b), which describes the binding energy per molecule at non-dilute conditions.

In the ideal case where the binding sites are homogeneous and receptors are dilute, the effects of intermolecular interactions in the activity coefficients vanish in Eq. (20). As a result, one obtains the classical Langmuir isotherm Langmuir (1918), and KL reduces to the Langmuir constant KL = exp(Δμ0/kBT), that is determined solely by the dilute binding affinity Δμ0.

In summary, starting from the general equilibrium conditions and incorporating conservation of mass and non-ideal interactions via activity coefficients, we have derived a general binding isotherm applicable to non-dilute conditions (19a). This result not only generalizes the theory of surface binding to non-dilute conditions recovering the classical Langmuir equation in the dilute limit, it also describes phase transitions at the binding surface leading to a heterogeneous distribution of bound molecules.

The non-dilute binding isotherm and surface phase separation

In this section, we investigate the intricate behaviors of non-dilute surface binding by varying bulk and surface compositions. Using Flory-Huggins free energy densities for the bulk-membrane system (Appendix 1), we assume unimolecular binding (α = β = 1) and examine the thermodynamic equilibrium satisfying the conditions of phase coexistence (16b)–(16d) and the generalized binding isotherm (20), we obtain with To align with physical constraints and simplify computational implementation, we set the basic parameter values as follows: ωPR = −1, χ = 2.5, n = nPR = nR = 1, ω = ωR = 0, χR-L = 0, and χPR-R = 0. Specifically, ωPR, the internal free energy coefficient, denotes the binding affinity of proteins to the membrane receptors. Setting ωPR = −1 reflects a moderate attractive interaction, consistent with observed protein-receptor binding in similar systems. The parameter χ = 2.5 ensures sufficient phase separation tendencies in the bulk, while n = nPR = nR = 1 assumes comparable molecular sizes between receptor-bound proteins, unbound receptors, and lipid molecules, simplifying the modeling framework. Finally, since the binary mixture of free receptors and lipids in the membrane does not undergo phase separation, direct receptor–lipid interactions are neglected. Accordingly, we set χR-L = 0. Due to strong binding in the case of 14mers receptors, free receptors are very dilute in the experiments, making the interaction parameter χPR-R an irrelevant parameter. Thus, we set χPR-R = 0 in this study. Finally, setting these interaction parameters to zero allows us to focus on the dominant interactions that determine the system’s binding and phase separation behavior.

We first consider the scenario where phase separation does not occur on the surface, as the attractive interactions between bound proteins (χPR-L = 2) are insufficient to induce surface phase separation. Under these assumptions, our analysis reveals that the binding fraction θ increases monotonically with the bulk protein volume fraction ϕ in the non-dilute case (Fig. 2a) This behavior qualitatively resembles the dilute case, which corresponds to a scenario where the total receptor concentration is very low ; see conservation law (3a). However, in the non-dilute case, the binding fraction θ is significantly higher than in the dilute case as the total receptor concentration increases. This enhancement is a direct consequence of non-dilute binding. At very high bulk protein volume fractions (ϕ), the bulk undergoes phase separation, further altering the system’s behavior.

Non-dilute surface binding behaviors with varying surface and bulk compositions.

(a) Binding fraction curves are shown for three different initial receptor area fractions: (solid line), (dashed line) and (dotted line), under the assumption of no surface phase separation (χPR-L = 2). (b) With a stronger interaction strength (χPR-L = 3.2) and a fixed initial receptor area fraction , the homogeneous surface mixture (black line) undergoes phase separation into two distinct phases: a protein-rich phase (red line) and a protein-poor phase (blue line) as a function of the intermediate bulk volume fraction, ϕ. Additionally, when the bulk protein volume fraction (ϕ) exceeds the saturation level, a 3D bulk condensate forms (blue region) that coexist with 2D surface condensates on the membrane. (c) The phase diagram illustrates that intermediate receptor area fractions and large bulk protein volume fractions (ϕ) promote surface phase separation on the membrane. The basic parameter values are: ωPR = −1, χ = 2.5, n = nPR = nR = 1, ω = ωR = 0, χR-L = 0, and χPR-R = 0.

We now examine the case where the bound proteins at the surface can undergo surface phase separation, forming a protein-rich and a protein-poor phase leading to a heterogeneous distribution of bound proteins PR at the surface (see Fig. 2b). A sufficiently large bulk protein concentration is needed to ensure enough proteins are bound to receptors at the surface. Moreover, interactions need to be strong enough (e.g., χPR-L = 3.2), e.g., the bound proteins attract each other sufficiently strongly.

The effects of both the bulk protein volume fraction ϕ and the total amount of receptors on the surface are summarized as a phase diagram in Fig. 2c for a fixed value of χPR-L = 3.2. It shows that surface phase separation occurs in an intermediate range of both bulk protein and total receptor area fraction . A too-large value of bulk protein volume fraction triggers phase separation in the bulk independent of the receptor and surface binding. Moreover, a too-large amount of receptors on the surface leads to a homogeneous state for volume fraction fractions below bulk phase separation since the mixing entropy on the surface dominates. Similarly, if there are too few receptors on the surface, the enthalpic gains related to the interaction among bound proteins are also outcompeted by the mixing entropy. This overall indicates a reentrant behavior of surface phase separation when varying the total receptor area fraction .

Effects of binding affinity and interactions on surface phase separation

We also investigated how the binding affinity (ωPR) and the interaction strength between membrane-bound proteins (χPR-L) affect surface phase separation. We distinguish two scenarios (Fig. 3a): The three-dimensional bulk is homogeneous, or the bulk is phase-separated.

Surface behavior with varying binding affinity and membrane-bound interaction strength.

(a) Schematic illustrating bulk-membrane coupled system with surface condensates formed in the membrane S without and with bulk phase separation, respectively. To compare phase behavior with and without bulk demixing, we suppress bulk phase separation by setting the protein–solvent interaction to χ = 0 and, for each binding affinity ωPR and membrane–complex interaction strength, compute the surface-binding curve (Fig. 2) to test for surface phase separation. We show the result in phase diagram (b). We then contrast this with a case of stronger bulk interactions (χ = 2.5) in (c). If protein interaction in the bulk is strong and thus bulk phase separation occurs. With bulk condensates partially/completely wet on top of the membrane, the surface phase separation occurs in a much larger parameter regime. This illustrates that bulk protein properties, like interaction strength, also regulate the surface phase behavior. The basic parameter values we set in this study are: , (b) χ = 0, (c) χ = 2.5. Here, csat represents the bulk volume fraction at the saturation level.

In the absence of bulk phase separation, as shown in Fig. 3b, we find that upon increasing the binding affinity (ωPR) and membrane-bound interaction strength (χPR-L) can induce surface phase separation on the membrane, shifting the system from a homogeneous state to a phase-separated one on the surface. This result underscores the importance of binding affinity and membrane-bound complex interactions as a driving force for surface phase separation.

In the presence of bulk phase separation, as shown in Fig. 3c, we observe a qualitatively similar behavior: strong binding affinity (more negative values of ωPR) and a larger membrane-bound interaction strength (χPR-L) can induce phase separation on the membrane surface. However, the domain of surface phase separation is much larger than the one without bulk phase separation (compare to Fig. 3b), indicating that the bulk interaction parameter χ affects phase separation on the membrane. Mechanistically, a larger (and positive) bulk protein interaction strength χ increases the local chemical potential of free proteins μ near the membrane. As a result, the binding equilibrium (16a) is changed with an enlarged chemical potential difference between unbound and bound states, making more proteins bind from the bulk to the membrane. This leads to the formation of more membrane-bound complexes, making surface phase separation occur over a much broader range of binding affinities and interaction strengths, as illustrated in Figure 3c. In summary, surface phase separation is collectively regulated by the bulk protein interaction strength (χ), the binding affinity (ωPR), and the interaction strength among membrane-bound proteins (χPR-L).

Having established the theoretical framework, we now apply it to extract mechanistic insight into tight junction formation, combining with in-vitro reconstitution experiments.

Mechanism in tight junction formation

Tight junctions are important membrane-associated structures for epithelial and endothelial barrier formation. How surface binding interplays with the condensation process of ZO1 on the plasma membrane at the cell-cell contact sites is not well understood. In this section, we will understand the mechanism for surface phase separation involved in the formation of tight junctions, combining our developed theoretical model and in-vitro reconstitution experiments.

At the cell-cell interface, tight junction receptors including the Claudin family, the CTX family (JAMs) and the MARVEL family (Occludin, Tricellulin, MARVELD3), form multiple trans- and cis-interactions to mediate cell adhesion (Fig. 4a). Genetic mutation in these trans- and cis-interactions has been shown to induce tight junction and epithelial barriers deficiency Kostrewa et al. (2001); Zhao et al. (2018b); Heinemann and Schuetz (2019). In this system, we consider ZO1 as the scaffold proteins in the bulk, Claudin2 C terminal (CLDN2-ICD, referred as 1-mer hereafter), which is one of the essential receptors for tight junction function, as the receptors anchored in the membrane. CDLN2-ICD were tagged with His tag and anchored to the membrane with DGS-NTA (Ni) lipid as a mimic of the membrane-associated receptors via His-Ni interaction (Fig. 4b). To mimic the receptor oligomerization happening at the cell-cell contact site, ySMF domain, which is known to form stable 14-mer Collins et al. (2003), was introduced to the CDLN2-ICD construct to get 14-meric receptors (referred as 14-mer hereafter). Scaffold proteins ZO1 were then added to the bulk. See Appendix 4 for methods and materials used in the in vitro experiments.

Illustration of tight junction initiation and in vitro reconstitution system.

(a) Schematic representation of tight junction initiation during cell-cell contact. Upon contact between Cell 1 and Cell 2, receptors (magenta) on the plasma membrane (PM) interact with scaffold proteins such as ZO1 (green) to form condensates that contribute to tight junction formation Beutel et al. (2019). (b) In vitro reconstitution system showing the binding dynamics of ZO1 proteins (green) with receptors (magenta) on a supported lipid bilayer. The His(10)-tagged receptors are anchored to DGS-NTA (Ni) lipid via His-Ni interaction and can bind and unbind with the ZO1 proteins in the bulk.

Now we use our theory for non-dilute binding to membrane-like surfaces and surface phase separation of bound proteins to model an in vitro reconstitution system for tight junction initiation.

We will show that the theory will be instrumental to characterize binding affinities at non-dilute conditions and elucidate the role of the oligomerization state of the receptor. As a crucial step in this analysis, we estimate the fundamental model parameter values for ZO1 protein binding to claudin receptors by integrating our non-dilute binding theory and experimental data (see Appendix 5 for details). These parameter values, which form the basis of subsequent modeling, are summarized in Table 1.

Receptor oligomerization facilitates surface phase transitions

We analyzed how the oligomerization state of the receptor affects the surface phase transition. In this study, we used two kinds of receptor oligomerization states: 1-mer and 14-mer. By fitting our model (16) to the experimental data (see Appendix 6) for both receptor states (Fig. 5c and 5d and Fig. 6c and 6d), we determined the changes in receptor binding affinity ωPR and self-interaction parameter χPR-L of membrane-bound ZO1 complexes. A key finding is that the binding affinity of the 14-meric receptor to ZO1 protein (ωPR = 3.83, Fig. 6) is much stronger compared to the monomeric receptor (ωPR = −1.09, Fig. 5). Furthermore, we found that the lower binding affinity of the 1-mer is dominantly caused by a reduced binding fraction, i.e., only a small fraction of 1mer receptors can bind to ZO1 (see Appendix 7 for details); a mechanism previously reported in Ref. Erlendsson et al. (2019). Moreover, from the fits, we obtained that the bound ZO1-14-mer complexes strongly attract each other, i.e., they have a positive self-interaction parameter (Fig. 6). While the fitted self-interaction of the ZO1-1-mer (χPR-L = −0.4) complex was lower compared to the 14-mer(χPR-L = 1.09), a wide range of self-interaction values provide similar good fits (blue shade in Appendix 6-fig. 1c), indicating that in the low-affinity binding regime of the 1-mer receptor self-interactions at the surface play a minor role (see Appendix 6-fig. 1c and its related discussion).

Surface stays homogeneous with 1-mer receptor anchored on the membrane.

(a) Fluorescence images of ZO1 and 1-mer at 1-mer receptor titrations, ranging from 19386 to 102689 μm−2. (b) Fluorescence images of ZO1 and 1-mer showing the effect of ZO1 titration (0 to 800 nM) on 1-mer binding. Scalar bars in the bottom panels represent 2 μm. (c) The linear relationship between membrane-bound ZO1 density and receptor density on the membrane. This is because, in this study, the binding is saturated for all the receptor densities. In this figure, the dots with error bars represent the experimental data (quantified from figure (a)), and the solid line depicts the theoretical prediction. (d) Curve of membrane-bound ZO1 density as a function of ZO1 bulk density compared between experimental data (points) and theoretical prediction (line). The plateau is obtained from the binding fraction limit for 1-mer, c = 0.014. See Appendix 7 for details. The data is a fit to the membrane-bound ZO1 density observed in panel (b). (e) Phase diagram illustrating the homogeneous surface state across a physiological range of receptor and ZO1 densities.

Surface condensates form with 14-mer receptor on the membrane

(a) Fluorescence images showing the distribution of 14-mer receptors and ZO1 across a range of receptor titrations, from 7 to 2744 μm−2. Surface phase separation occurs at intermediate receptor densities. Too few or too many receptors result in a homogeneous membrane phase: a dilute phase at low receptor densities and a dense phase at high receptor densities. (b) Fluorescence images of 14-mer receptors and ZO1 as a function of increasing ZO1 bulk concentration, from 4 to 14000 nM, demonstrating the emergence of phase-separated domains. Since receptor density remains constant during this titration, the area of surface condensates does not change with increasing ZO1 bulk concentration. Once the ZO1 bulk concentration exceeds the saturation level, bulk condensates form in the presence of surface condensates (refer to Fig. 3a for illustration). Scalar bars in the bottom panels represent 2 μm. (c) Membrane-bound ZO1 density as a function of receptor density on the membrane, showing how receptor density influences surface phase behavior. In this figure, the dots with error bars represent the experimental data (quantified from figure (a)), and the solid line depicts the theoretical prediction. (d) Membrane-bound ZO1 density dependence on ZO1 bulk concentration, comparing experimental data with theoretical predictions. The data is a fit to the membrane-bound ZO1 density observed in figure (b), revealing a transition from a homogeneous surface phase to a two-phase coexisting state (protein-poor and protein-rich phases). (e) Phase diagram with respect to the receptor density on the membrane and Bulk ZO1 density, indicating that 14-mer receptor promotes surface phase separation.

In summary, our results suggest that receptor oligomerization promotes surface phase separation through two synergistic mechanisms. First, increased binding affinity leads to a higher concentration of ZO1 on the membrane. Second, the strong self-interaction of the protein-receptor complex lowers the saturation concentration required for surface phase separation.

Using the parameters obtained from the fits, we determine the phase diagram for the monomeric (Fig. 5e) and 14-meric receptors (Fig. 6e). These phase diagrams show that there is no surface phase transition for monomeric receptors even for higher ZO1 concentrations, while there is a large region of surface phase separation for the 14-meric receptors.

Based on the agreements between experiments and our thermodynamic model, we conclude that a surface phase transition drives the formation of ZO1-rich surface condensates on membrane-like surfaces. The surface phase transition is controlled by the receptor’s oligomerization state, suggesting cell adhesion-induced oligomerization could act as the key switch for ZO1 surface condensation at cell-cell contacts.

Spatial pattern formation of ZO1 surface condensates

When proteins bound to the surface undergo phase separation, multiple condensates form and exhibit a slow ripening kinetics. Theoretical results on surface binding, surface condensates growth and ripening can be obtained from numerically solving the dynamic equations (24) in a three-dimensional bulk volume V with surface S (Figure 7(a), see Appendix 2 and Appendix 3 for details). Both theoretical predictions and reconstitution experiments reveal an initial stage where surface condensates form and grow rapidly, followed by a prolonged ripening phase characterized by slow kinetics (see Appendix 8, Fig. 1). During this ripening stage, the average condensate areas predicted by the theory align well with experimental observations. Importantly, over a time course of approximately 15 minutes, the amount of bound proteins remains nearly constant as the surface condensates ripen, with only a slight reduction in the number of condensates. To further validate the model, we compared the condensate patterns from experiments and theory at time t = 15 min for different receptor titration levels ranging from 25 to 3800 PN/μm2 and ZO1 titration levels from 4 to 100 nM (Fig. 7(b,c)). The patterns show quantitative agreement between theory and experiments, providing strong evidence that the theoretical framework effectively cap-tures the non-dilute surface binding mechanisms and the condensate formation process on the surface. Furthermore, we observed that the receptor density on the membrane significantly influences the spatial patterns of surface condensates. At lower receptor densities, condensates tend to be sparse and less connected, while higher receptor densities promote larger and more interconnected condensates, consistent with predictions from the theoretical model (Fig. 7(b,c)).

Snapshots of ZO1 surface condensates from simulation (top) and experiments (bottom) when titrating the 14-mer receptor and ZO1 bulk concentration.

(a) Schematic of the system in our theoretical model, with 3D bulk volume V and membrane surface S, where surface condensates formed via surface phase separation. (b) Snapshots of ZO1 from simulation (top row) and experimental observations (bottom) under varying receptor titration levels, from 25 to 3800 μm−2, at 15 mins. (c) Snapshots of ZO1 from simulation (top row) and experimental observations under varying ZO1 concentration levels, from 4 to 100 nM, at 15 min. The experimental results match well with the theoretical predictions from perspectives of both condensate pattern and condensate size. Scale bars in the bottom panels represent 2 μm. we set the diffusivity coefficient value as Di = 0.01 μm2s−1, i = P, P R, R. The gradient coefficients κ = κPR = κR = 2 × 10−5μm−1.

Conclusion

In this study, we developed a comprehensive theoretical framework grounded in thermodynamics to describe how soluble scaffold proteins bind lipid membranes through interactions with membrane-embedded receptor proteins. The framework derives a generalized binding isotherm, based on the thermodynamic theory for non-dilute surface binding. This isotherm extends classical adsorption models by accounting for the non-dilute nature of membrane-bound complexes and their interactions. Compared to the classical Langmuir binding isotherm, our theory incorporates interactions among membrane-bound molecules, and accounts explicitly for membrane binding heterogeneity. Consequently, our model is able to capture complex processes such as surface phase separation, which are beyond the descriptive capabilities of classical dilute binding models. Additionally, a kinetic model was derived from non-equilibrium thermodynamics to investigate spatiotemporal patterns of surface condensates.

Using this framework, we explored the behaviors of non-dilute surface binding under varying bulk and surface compositions, binding affinities, and interaction strengths between membrane-bound complexes. The model also enabled us to simulate and analyze surface phase transitions in an in vitro reconstitution system for tight junction initiation. A critical insight from this work is the role of receptor oligomerization in facilitating surface phase transitions, highlighting its importance in membrane-associated condensation phenomena.

Our findings demonstrate that receptor density significantly influences the spatial patterns of surface condensates. At lower receptor densities, condensates are sparse and less connected, whereas higher receptor densities promote larger and more interconnected structures. Moreover, the patterns observed in simulations show quantitative agreement with experimental results, providing strong evidence that the theoretical framework accurately captures the mechanisms of non-dilute surface binding and condensate formation processes, which the dilute model fails to reproduce.

On the membrane with 1-mer receptors, the distribution of membrane-bound proteins remains homogeneous regardless of the bulk ZO1 concentration. The binding isotherm could also be well described by a dilute binding theory, even at non-dilute ZO1 concentrations(see Fig. 5). The reason is that our fits show that the interactions between protein–receptor complexes play only a minor role in this case. In contrast, for 14-mer receptors, surface phase separation emerges at concentrations far below the bulk saturation concentration (see Fig. 6), reflecting both a stronger binding affinity and enhanced intermolecular interactions among protein–receptor complexes on the membrane. Such behavior cannot be captured by the dilute binding isotherm. Therefore, a non-dilute binding theory is essential in our study to account for these complex surface phase behaviors.

In summary, this work bridges theoretical modeling and experimental validation, advancing our understanding of non-dilute surface binding and phase separation on biological membranes. However, it is worth noting that the current theoretical framework has also been applied in limiting cases. For example, in the present model, we looked at a flat membrane, thereby neglecting the effects of membrane curvature, which can significantly influence protein binding and likely also wetting transitions of bimolecular condensates Rascón et al. (2018); Lipowsky (2025). Moreover, we focused on the behavior at and close to equilibrium, and neglected fluxes related to dissipation at system boundaries, such as the triple line. Thus, a richer phenomenology is expected when ac-counting for far-from-equilibrium dynamics of surface phase separation. Addressing these richer scenarios will be crucial for applying our framework to in vivo systems of biomolecular complexes on cellular membrane surfaces.